SSA based global code motion of identical computations

### 2017-04-16 acmart-pacmpl-template.tex
legality

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. We have also open
sourced the implementation for easy verification. The main contributions of this
paper are:
\begin{itemize}
\item a new optimistic algorithm to move computations out of branches,
\item a cost model to reduce interference and reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  SSA based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

\subsection{Finding candidates to move}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution to value number loads is to hash the address from
where the value is to be loaded. For stores, we value number the address as well
as the value to be stored at that address.

Another limitation of the current \GVN{} implementation in \LLVM{} is that the
instructions dependent on the loads will not get numbered correctly, and so
after hoisting all candidates we need to rerun the \GVN{} analysis in order to
discover new candidates now available after having hoisted load instructions.
This limitation should be addressed in a new implementation of the \GVN{} based on
\MemorySSA{}, that would better account for equivalent loads and their dependent
instructions.

@@

Once the instructions have been classified in equivalence classes, we compute
for each group of equivalent instructions a point in the program that is both
legal and profitable for the instructions to be moved to.

\subsection{Legality check}
\label{subsec:legality}
Since the equality of candidates is purely based on the value they compute, we
need to establish if hoisting them to a common dominator or sinking them to a
common post-dominator would be feasible. Once a common dominator
(post-dominator) is found, we check whether all the use-operands of the set of
instructions are available at that position. In some cases when the operands are
not available, it is possible to reinstantiate (remateralize) the use-operands,
thus passing the legality check.

Subsequently, it is checked that the side-effects of the computations does not
intersect with any side-effects between the instructions to be hoisted/sunk and
their hoisting/sinking point.

Moreover, moving memory operations is tricky on paths which have indirect
branch targets e.g., landing pad, case statements, goto labels etc., because it
becomes difficult to prove safety checks. In our current implementation we
discard candidates on those paths.

In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point of all the instructions is either too far away, or
not legally possible. In these cases, it is still possible to `partially' hoist
a subset of instructions by splitting the set of candidates and finding a closer
hoisting point for each subset. For more details see
Section~\ref{subsec:partition}.

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available (or can be
made available by rematerialization), the scalar computation can be hoisted.

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case we discard the
candidate.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads. We check that the operands of the store instruction are available at the
hoisting point, that there are no aliasing loads or store along the path from
the current position to the hoisting point.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and most of the time,
without further information, calls have to be classified as writing to memory,
that is the most restrictive form.  Each category of call instructions will be
handled as described for scalar, load, and store instructions.

\section{Profitability check (Cost models)}
\label{sec:cost-models}
After the legality checks have passed, we check for profitability of hoisting.
That takes into account the impact code-hoisting would have on various
parameters that affect runtime performance e.g., impact on live-range, gain in
the code size. Since this is mostly a code-size optimization pass, the goal is
to not regress in performance across popular benchmarks at the same time reduce
code size as much as possible. Following are the cost models which are
implemented:

@@

\subsection{Optimistic code motion algorithm}
\label{subsec:optimistic}
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure. In those cases, this algorithmn would partition the list of
identical instructions into subsets which can be partially hoisted to their
respective common dominators. However, it should be noted that we only hoist
very busy expressions to avoid checking for undefined behaviors resulting because
of introducing extra computation in an execution path.

@@

For sinking however, higher ranked expressions would be sunk first. And it would
be illegal to sink higher ranked identical expressions if they are not
anticipable in the common post-dominator. For example:

In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
on i1 and i2 respectively which are also in their original basic blocks, c1 and
c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
safely be sunk because their operands are readily available at the sink point
i.e., B3. It should also be noted that, just because the expressions are
identical and operands are available it still requires a unique post-dominator
PHI to use the exact same values to be legally sinkable.

The partition algorithm is described as follows:
% TODO: Explain why all the identical computation cannot be hoisted.

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and for
some reason they cannot be hoisted together at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
B1 and those in B5, B6 to be hoisted at B2.

@@

Currently, we rerun the algorithm until there are no more instructions left to
be hoisted. This is not the most optimal approach and results in expensive
analyses to be recomputed. This can be improved by ranking the computations
\cite{rosen1988global}. Also \GVN{}-hoist runs very early in the pass pipeline, it
will be good to evaluate the codesize/performance impact when it is run in sync
with \GVN{}-\PRE{} just like \GCC{} does.

>>>>

There have been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using ILP which results in prohibitively high compile time. To the best
of our knowledge we have not found any reference which explored code-motion of
identical computation in as much detail as we have done. A part of our
implementation (aggressive code hoisting) is already merged in LLVM trunk,
however, a general implementation of code-motion of identical expressions is
still missing from GCC and LLVM trunk. The main contributions of this paper are:
\begin{itemize}
\item a new optimistic algorithm to move computations out of branches,
\item a cost model to reduce interference and reduce spills,
\item a technique to maximize hoisting in an optimistic approach by partitioning
  the list of potential candidates sorted by their DFS visit number,
\item experimental evaluation of our implementation in \LLVM{} which combines
  SSA based liveness analysis, and ranking of expressions to move very busy
  expressions in order to reduce code-size (and improve performance in some
  cases).
\end{itemize}

@@

\subsection{Finding candidates to move}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}. We
could computed available and anticipable sets as computed by GCC's code-hoisting
but that would be a lot of data structures to maintain at each basic block
level. For GCC it makes sense because their code-hoisting is integrated with
GVN-PRE which already has those data structures available.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution to value number loads is to hash the address from
where the value is to be loaded. For stores, we value number the address as well
as the value to be stored at that address. Another limitation of the current
\GVN{} implementation in \LLVM{} is that the instructions dependent on the loads
will not get numbered correctly, and so after hoisting all candidates we need to
rerun the \GVN{} analysis in order to discover new candidates now available
after having hoisted load instructions.  This limitation should be addressed in
a new implementation of the \GVN{} based on \MemorySSA{}, that would better
account for equivalent loads and their dependent instructions.

@@

Once the instructions have been classified into equivalence classes, we compute
for each group of equivalent instructions, a point in the program that is both
legal and profitable for the instructions to be moved to.

\subsection{Legality check}
\label{subsec:legality}
Since the equality of candidates is purely based on the value numbers, we also
need to establish if hoisting them to a common dominator or sinking them to a
common post-dominator would be legal. Once a common dominator (post-dominator)
is found, we check whether all the use-operands of the set of instructions are
available at that position. It is possible reinstantiate (remateralize) the
use-operands in some cases when the operands are not available and make it legal
to move the instruction.

Subsequently, it is checked that the side-effects of the computations (if any)
do not intersect with any side-effects between the instructions to be
hoisted/sunk and their hoisting/sinking point. It is also necessary to check if
there are indirect branch targets e.g., landing pad, case statements, goto
labels etc., along the path because it becomes difficult to prove safety checks
in those cases. In our current implementation we discard candidates on those
paths.

Code-motion basically consists of two parts i.e., hoisting and sinking. The
unique part about this algorithm is that it only moves identical
computations. An immediate guarantee of this approach is gain in code size (the
final executable may be of larger size because of more inlining). The algorithm
prefers hoisting to sinking. If the dependency of a hoistable candidate is in
the same basic block as the candidate, then the dependency must also be
hoistable otherwise hoisting will be illegal or would require a complicated code
generation to make it legal. The current algorithm discards cases if the
dependency is neither hoistable nor rematerializable.

\subsubsection{Legality of hoisting scalars}
Scalars are the easiest to hoist because we do not have to analyze them for
aliasing memory references. As long as all the operands are available (or can be
made available by rematerialization), the scalar computations can be hoisted.

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the addresss if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case we discard the
candidate. We use the MemorySSA infrastructure of \LLVM{} to iterate on the
use-def chains available for memory references.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the MemorySSA of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, that there are no aliasing
loads or store along the path from the current position to the hoisting point.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and most of the time,
without further information, calls have to be classified as writing to memory,
that is the most restrictive form.  Each category of call instructions is
handled as described for scalar, load, and store instructions.

\subsubsection{Legality of sinking expressions}
One might wonder that if a set of expressions compute the same value then why
not just hoist them. The problem is, sometimes, hoisting is not upward-safe
\ref{click1995global} e.g., if the expressions are in a landing pad
etc. Sometimes sinking may reduce the register pressure e.g., when the use
operands are not kills.

%TODO: Explain why all the identical computation cannot be hoisted.
For sinking, higher ranked expressions would be sunk first. And it would be
illegal to sink higher ranked identical expressions if they are not anticipable
in the common post-dominator. For example:

In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
on i1 and i2 respectively which are also in their original basic blocks, c1 and
c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
safely be sunk because their operands are readily available at the sink point
i.e., B3. It should also be noted that, just because the expressions are
identical and operands are available, it still requires a unique post-dominating
PHI to use the exact same values to be legally sinkable.

A general global scheduling algorithm also requires checks for undefinedness
when introducing a new computation along a path. Since only very busy
expressions are moved in the current implementation, there is no need to check
for undefinedness resulting due to movement of instruction. This simplifies the
implementation.

\subsubsection{Partitioning the list of hoistable candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach, described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point of all the instructions is either too far away, or
not legally possible. In these cases, it is still possible to `partially' hoist
a subset of instructions by splitting the set of candidates and finding a closer
hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and for
some reason they cannot be hoisted together at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in B3, B4 to be hoisted at
B1 and those in B5, B6 to be hoisted at B2.

\section{Profitability check (Cost models)}
\label{sec:cost-models}
After the legality checks have passed, we check if a code-motion is profitable.
That takes into account the impact code-motion would have on various parameters
that might affect runtime performance e.g., impact on live-range, gain in the
code size. Since this is mostly a code-size optimization pass, the goal is to
not regress in performance across popular benchmarks at the same time reduce
code size as much as possible. Following are the cost models which are
implemented:

@@

\subsection{Optimistic code motion algorithm}
\label{subsec:optimistic}
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure. In those cases, this algorithmn would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators. However,
it should be noted that we only hoist very busy expressions to avoid checking
for undefined behaviors resulting because of introducing extra computation in an
execution path.

@@

@@

In general it is a good idea to start with lower ranked expression first such
that maximum hoisting can happen in one iteration, however, current
implementation does not rank the expressions and iteratively finds a fixed point
when no more candidates are available. Even this implementation converges
quickly and no significant compile time regression have been observed because of
code hoisting pass. This is not the most optimal approach and results in
multiple data structures to be recomputed. This can be improved by ranking the
computations \cite{rosen1988global}. Also \GVN{}-hoist runs very early in the
pass pipeline, it will be good to evaluate the codesize/performance impact when
it is run in sync with \GVN{}-\PRE{} just like \GCC{} does.
